---
title: "preprocess"
author: "markrobinsonuzh"
date: "2019-01-28"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: console
---

## Introduction

Before any of this, you will need to download the processed data from GEO at accession [GSE96583](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE96583). The file `GSE96583_RAW.tar` was unwrapped into the `GSE96583_RAW` folder and then files in there were moved to the `/data/` folder of the workflow here. The files `GSE96583_batch2.genes.tsv.gz` and `GSE96583_batch2.total.tsne.df.tsv.gz` were also individually downloaded and placed in the `/data/` directory.

Assuming you have the Unix tool `tree` installed, you should have the following set of files in the `data` directory (although only those specified below are strictly needed):

```{r}
twee <- system("tree data", intern = TRUE)
cat(paste(twee, collapse = "\n"))
```

## Load packages

```{r}
# load packages
suppressPackageStartupMessages({
    library(dplyr)
    library(SingleCellExperiment)
})
```

## Load data into Bioconductor objects

```{r}
# load data
rd <- read.table("data/GSE96583_batch2.genes.tsv.gz", 
                 sep="\t", header=FALSE, stringsAsFactors=FALSE,
                 row.names=1, col.names=c("", "feature"))
rd$gene_id <- rownames(rd)
rd$symbol <- rd$feature
rd$feature <- NULL

cd <- read.table("data/GSE96583_batch2.total.tsne.df.tsv.gz", 
                 sep="\t", header=TRUE)

mtx <- file.path("data", c("GSM2560248_2.1.mtx.gz",
                           "GSM2560249_2.2.mtx.gz"))
x <- lapply(mtx, Matrix::readMM)
x <- do.call(cbind, x)
dimnames(x) <- lapply(list(rd, cd), rownames)
dim(x)

# remove undetected genes
x <- x[Matrix::rowSums(x > 0) > 0, ]
rd <- rd[rownames(x), ,drop=FALSE]

rownames(x) <- rownames(rd) <- with(rd, paste0(gene_id, "_", symbol))

# remove multiplets & unassigned cells
cd <- data.frame(cd, i=seq_len(ncol(x))) %>%
    filter(multiplets == "singlet", !is.na(cell))
rownames(cd) <- colnames(x)[cd$i]
cd <- select(cd, -"i")
x <- x[, rownames(cd)]
dim(x)

# construct rowData & colData
sample_ids <- factor(with(cd, paste0(stim, ind)))
cd <- data.frame(sample_id = sample_ids,
                 group_id=cd$stim, cluster_id=cd$cell,
                 row.names=rownames(cd))

# construct metadata
m <- match(levels(sample_ids), sample_ids)
ei <- data.frame(sample_id = levels(sample_ids),
                 group_id = cd$group_id[m])

# construct SingleCellExperiment
sce <- SingleCellExperiment(
    assays = list(counts = x),
    rowData = rd, colData = cd,
    metadata = list(experiment_info = ei))
sce <- scater::normalize(sce)

saveRDS(sce, "output/GSE96583_sce.rds")

# quick summary of assigned cells
with(colData(sce), table(sample_id, cluster_id))
```
